The Ky Fan Norms and Beyond: Dual Norms and Combinations for Matrix Optimization

### 2025-10-10 slides/neon_slides.tex
longer presentation done

@@

% Title
\title[The Ky Fan Norms and Beyond]{The Ky Fan Norms and Beyond: Dual Norms and Combinations for Matrix Optimization}
\author[A. Kravatskiy et al.]{Alexey Kravatskiy\inst{1} \and Ivan Kozyrev\inst{1} \and Nikolai Kozlov\inst{1} \and Alexander Vinogradov\inst{1} \and Daniil Merkulov\inst{1,2,3,4} \and Ivan Oseledets\inst{5,2}}
\institute[]{%
\inst{1} MIPT \and
\inst{2} Skoltech \and
\inst{3} HSE \and
\inst{4} SberAI \and
\inst{5} AIRI
}
\date{\today}

@@

    \vspace{0.3em}
    \footnotesize Dual balls for F-Muon and F-Neon across \(\alpha\) in a 2D singular-value space
    \end{frame}
%---------------------------------------
\begin{frame}{F-Muon on CIFAR-10 airbench}
    \begin{columns}[T,totalwidth=\textwidth]
      \begin{column}{0.48\textwidth}
        \includegraphics[width=\linewidth]{muon_tuned_diff_alpha.pdf}
        \centering
        \scriptsize Params tuned for Muon \citep{cifar2023airbench}
      \end{column}
      \begin{column}{0.48\textwidth}
        \includegraphics[width=\linewidth]{fmuon_tuned_diff_alpha.pdf}
        \centering
        \scriptsize Params tuned for F-Muon
      \end{column}
    \end{columns}
    \vspace{0.6em}
    \centering

    \footnotesize F-Muon with \(\alpha\approx 0.5\) matches Muon after tuning
  \end{frame}

@@

    \faQuestionCircle \space How should we interpret cases with $\alpha > 1$?
  \end{figure}
\end{frame}
%-----------------------------------------------------------------------------------------------
\begin{frame}{NanoGPT speedrun}
    \begin{itemize}
      \item Setup from \citet{modded_nanogpt_2024}: 1750 iterations and cross-entropy loss lower than 3.28
      \item Muon: {\tt lr=0.05, momentum=0.95}, 3.279
      \item F-Muon with \(\alpha=0.5\): {\tt lr=0.07, momentum=0.95}, 3.281
    \end{itemize}
\end{frame}
%----------------------------------------
\begin{frame}{Trust-region bounds (L-smooth)}
  \begin{itemize}
    \item Deterministic: choose \(\eta=\cO(\epsilon/L)\), \(K=\cO(L\Delta_0/\epsilon^2)\) \citep{kovalev2025understanding}.
    \item Stochastic (with variance \(\sigma^2\), link between norms \(\rho\)): detailed schedules for \(\eta,\alpha,K\).
    \item Asymptotically, norms are proportional for large random matrices (Marchenko–Pastur), aligning guarantees across choices.
  \end{itemize}
\end{frame}

%----------------------------------------
\begin{frame}{Experiment: Random linear least squares}
  \begin{columns}[T,totalwidth=\textwidth]
    \begin{column}{0.48\textwidth}
      \includegraphics[width=\linewidth]{simple_lls/loss_vs_iteration_50x50.pdf}
      \centering
      \scriptsize Loss vs iteration
    \end{column}
    \begin{column}{0.48\textwidth}
      \includegraphics[width=\linewidth]{simple_lls/gradient_spectral_norm_vs_iteration_50x50.pdf}
      \centering
      \scriptsize Spectral norm of gradient vs iteration
    \end{column}
  \end{columns}
  \vspace{0.3em}
  \footnotesize Setup informed by \citet{kovalev2025understanding}.
\end{frame}

%----------------------------------------
\begin{frame}{Ky Fan geometry}
  \begin{figure}[t]
    \centering
    \includegraphics[width=0.45\linewidth]{KyFan.pdf}\hfill
    \includegraphics[width=0.45\linewidth]{KyFanDual.pdf}
    \caption{Ky Fan 2-norm ball and its dual (singular-value space).}
  \end{figure}
\end{frame}

%----------------------------------------
\begin{frame}{Conclusions}
  \begin{itemize}
    \item Fanions unify Muon/Neon via Ky Fan norms; F-Fanions add robustness via Frobenius mixing.
    \item TRLan enables efficient k-rank updates; practical speed benefits over power/RSVD.
    \item Empirically, F-Muon remains competitive on CIFAR-10 and NanoGPT.
  \end{itemize}
\end{frame}

>>>>

\usepackage{environ}
\newcounter{aequation}
\NewEnviron{aequation}{\refstepcounter{aequation}$$\BODY\eqno{\text{(A\theaequation)}}$$}
\crefname{aequation}{assumption}{assumptions}
\creflabelformat{aequation}{(#2A#1#3)}

@@

% Title
\title[The Ky Fan Norms and Beyond]{The Ky Fan Norms and Beyond: Dual Norms and Combinations for Matrix Optimization}
\author[A. Kravatskiy et al.]{%
  Alexey Kravatskiy\inst{1} \and
  Ivan Kozyrev\inst{1} \and
  Nikolai Kozlov\inst{1} \and
  Alexander Vinogradov\inst{1} \and
  Daniil Merkulov\inst{1,2,3,4} \and
  Ivan Oseledets\inst{5,2}\\[0.3em]
}
\institute[]{%
\inst{1} MIPT \hspace{1em}%
\inst{2} Skoltech \hspace{1em}%
\inst{3} HSE \hspace{1em}%
\inst{4} AI4Science \hspace{1em}%
\inst{5} AIRI
}
\date{\today}

@@

    \vspace{0.3em}
    \footnotesize Dual balls for F-Muon and F-Neon across \(\alpha\) in a 2D singular-value space
    \end{frame}
%---------------------------------------
\begin{frame}{F-Muon on CIFAR-10 airbench}
    \begin{columns}[T,totalwidth=\textwidth]
      \begin{column}{0.48\textwidth}
        \includegraphics[width=\linewidth]{muon_tuned_diff_alpha.pdf}
        \centering
        \scriptsize With params tuned for Muon, as in \citep{cifar2023airbench}
      \end{column}
      \begin{column}{0.48\textwidth}
        \includegraphics[width=\linewidth]{fmuon_tuned_diff_alpha.pdf}
        \centering
        \scriptsize With params tuned for F-Muon
      \end{column}
    \end{columns}
    \vspace{0.6em}
    \centering

    \footnotesize F-Muon with \(\alpha = 0.5\) matches Muon after tuning

    \faQuestionCircle \space How should we interpret cases with $\alpha > 1$?
  \end{frame}

@@

  \end{figure}
\end{frame}
%-----------------------------------------------------------------------------------------------
\begin{frame}{NanoGPT speedrun}
  Setup from \citet{modded_nanogpt_2024}: 1750 iterations and cross-entropy loss lower than 3.28
    \begin{itemize}
      \item Muon: {\tt lr=0.05, momentum=0.95}, final loss = 3.279
      \item F-Muon with \(\alpha=0.5\): {\tt lr=0.07, momentum=0.95}, final loss = 3.281
      \item NSGD: {\tt lr=0.07, momentum=0.96}, final loss = 3.4651
      \item F-Muon with $\alpha=0.5$: {\tt lr=0.07, momentum=0.96}, final loss = 3.2824!
    \end{itemize}
\end{frame}
%----------------------------------------------------------------------------------------------
\begin{frame}{Formal assumptions}
  L-smoothness:
  \begin{aequation}\label{eq:L}
    \norm{\nabla f(\mX) - \nabla f(\mX')}^\dagger \leq L \norm{\mX-\mX'}
    \quad\text{for all}\;
    \mX, \mX' \in \Rmn,
    \end{aequation}

  Bounded variance:
    \begin{aequation}\label{eq:variance}
      \Ed{\xi \sim \cD}{g(\mX;\xi)} = \nabla f(\mX)
      \quad\text{and}\quad
      \Ed{\xi \sim \cD}{\sqnf{g(\mX;\xi) - \nabla f(\mX)}} \leq \sigma^2
      \quad\text{for all}\;
      \mX \in \Rmn,
      \end{aequation}

  Bounded norm:
      \begin{aequation}\label{eq:norm}
        \norm{\mX}^\dagger \leq \rho\cdot\normf{\mX}
        \quad\text{for all}\;
        \mX \in \Rmn,
        \end{aequation}

  From Marchenko-Pastur law, $\normf{\cdot} \sim \frac{\sqrt{n}}{2}\norms{\cdot}$ and $\normn{\cdot} \sim \frac{n}{2}\norms{\cdot}$ for large $n$.

\end{frame}
\begin{frame}{Bounds from \citep{kovalev2025understanding}}
  \begin{lemma}\label{lemma:stoch_tr}
    To reach the precision $\E{\min_{k=1\ldots K} \norm{\nabla f(\mX_k)}^\dagger} \leq \epsilon$ under the assumptions \Cref{eq:L}, \Cref{eq:variance}, \Cref{eq:norm}, it is sufficient to choose the parameters as follows:
    \begin{align}
        \eta &= \cO\left(\min\left\{\frac{\epsilon}{L}, \frac{\epsilon^3}{\rho^2\sigma^2L}\right\}\right),
        \qquad
        \alpha = \cO\left(\min\left\{1, \frac{\epsilon^2}{\rho^2\sigma^2}\right\}\right),
        \\
        \label{eq:str_K_nonconvex}
        K &= \cO\left(\max\left\{
            \frac{\rho\sigma}{\epsilon},
            \frac{\rho^3\sigma^3}{\epsilon^3},
            \frac{L\Delta_0}{\epsilon^2},
            \frac{L\Delta_0\rho^2\sigma^2}{\epsilon^4}
        \right\}\right).
    \end{align}
    \end{lemma}

\end{frame}
%----------------------------------------
\begin{frame}{Random linear least squares: Loss and the Spectral Norm of the gradient}
  \begin{columns}[T,totalwidth=\textwidth]
    \begin{column}{0.48\textwidth}
      \includegraphics[width=\linewidth]{simple_lls/loss_vs_iteration_50x50.pdf}
      \centering
      \scriptsize Loss vs iteration
    \end{column}
    \begin{column}{0.48\textwidth}
      \includegraphics[width=\linewidth]{simple_lls/gradient_spectral_norm_vs_iteration_50x50.pdf}
      \centering
      \scriptsize Spectral norm of gradient vs iteration
    \end{column}
  \end{columns}
  \vspace{0.3em}
  \centering
  \footnotesize Linear least squares problem for a 50 x 50 matrix. Common {\tt lr=0.01}.
\end{frame}

%----------------------------------------
\begin{frame}{Geometric intuition: Ky Fan norms}
  \centering
  \begin{minipage}{0.46\textwidth}
      \includegraphics[height=0.75\textheight,keepaspectratio]{KyFan.pdf}
  \end{minipage}\hfill
  \begin{minipage}{0.46\textwidth}
      \includegraphics[height=0.75\textheight,keepaspectratio]{KyFanDual.pdf}
  \end{minipage}

  \vspace{0.3em}
  \footnotesize Ky Fan rank-2 ball and its dual in a 3D singular-value space
  \end{frame}
%----------------------------------------
\begin{frame}{Conclusions}
  \begin{itemize}
    \item Muon and NSGD are special cases of F-Fanions
    \item Mixture of norm-based updates is a norm-based update!
    \item No existing bounds describe the superiority of the spectral norm
  \end{itemize}

  \vspace{2em}
  \textbf{For future experiments}: Exploration of matrix and vector algorithms correspondence
\end{frame}
